---
title: "DATA607 Project 1 - Elo "
author: "David Simbandumwe"
output:
  pdf_document: default
  html_document:
    includes:
      in_header: header.html
    css: lab.css
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float: true
editor_options: 
  
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readtext)
```


# Summary
This project starts with a text file of chess tournament results. This file is processed to develop an R Markown file that generates a .CSV file that includes:
+ Player’s Name, 
+ Player’s State, 
+ Total Number of Points, 
+ Player’s Pre-Rating, 
+ Average Pre Chess Rating of Opponent



# Load Files

```{r, read file}

 eloTxt <- read_delim( file = "https://raw.githubusercontent.com/dsimband/DATA607/main/Project1/tournamentinfo.txt",              
          delim = " ",
          col_names = c("rec"),
          trim_ws = TRUE)


# clean data frame remove extra lines
eloTxt <- eloTxt %>% 
  filter(!str_detect(rec,"\\--"))

# remove heading rows
eloTxt <- eloTxt %>% slice(-c(1,2))


```




# Process Odd Rows

```{r, process odd rows}

eloTxt_odd <- eloTxt %>% filter(row_number() %% 2 == 1)


eloTxt_odd <- eloTxt_odd %>% 
      separate(rec, 
            into = c("id", "name" , "score" , "r1", "r2", "r3", "r4", "r5", "r6", "r7","x" ),
            extra = "merge", 
            fill = "left",
            convert = TRUE,
            sep = "\\|"
            ) %>% gather(
              "r1","r2","r3","r4","r5","r6","r7", key = "round", value = "op_data"
      )%>% 
      separate(op_data, 
            into = c("r_result", "op_id"),
            convert = TRUE
      ) %>% 
      drop_na(op_id)

```



# Process Even Rows


```{r, even rows}

eloTxt_even <- eloTxt %>% filter(row_number() %% 2 == 0)


eloTxt_even <- eloTxt_even %>% 
        separate(rec, 
              into = c("state", "score_data" , "x" ),
              extra = "merge", 
              fill = "left",
              convert = TRUE,
              sep = "\\|") %>% 
        mutate(id = row_number()) %>% 
        separate(score_data, 
              into = c("x1", "x2", "x3", "pre_rating", "post_rating"),
              extra = "merge", 
              fill = "left",
              convert = TRUE) %>% 
        separate(pre_rating, 
              into = c("pre_rating", "x4"),
              extra = "merge", 
              fill = "right",
              convert = TRUE,
              sep = "P") %>% 
        separate(post_rating, 
              into = c("post_rating", "x5"),
              extra = "merge", 
              fill = "right",
              convert = TRUE,
              sep = "P") %>% 
        select(id,state,pre_rating, post_rating)


```



# Join All Rows

```{r, join odd and even rows}

eloTxt_comb  <- eloTxt_odd %>% 
  select (id, name, score, round, r_result, op_id) %>%
  left_join(eloTxt_even, by = "id") %>%
  left_join(eloTxt_even, by = c("op_id" = "id"), suffix = c("", ".op"))


```





# Calculate 

```{r, calculate elo for wins and losses}

win_df <- eloTxt_comb %>%
  group_by(id,r_result) %>%
  filter(str_detect(r_result, "W")) %>%
  mutate(
    win_elo = sum(pre_rating.op) + n()*400,
    win_num = n()
  ) %>%
  select(id,r_result,win_elo, win_num) %>%
  distinct()

loss_df <- eloTxt_comb %>% group_by(id,r_result) %>%
  filter(str_detect(r_result, "L")) %>%
  mutate(
    loss_elo = sum(pre_rating.op) - n()*400,
    loss_num = n()
  ) %>%
  select(id,r_result,loss_elo, loss_num) %>%
  distinct()

```



# create the summary table

```{r, summary table}

eloFinal <- eloTxt_comb %>% 
  select (id, name, state, score, round, r_result, op_id, pre_rating, post_rating, pre_rating.op, post_rating.op) %>%
  left_join(win_df, by = "id") %>%
  left_join(loss_df, by = "id") %>%
  replace_na(list(win_elo = 0, win_num = 0, loss_elo = 0, loss_num = 0 ))

# caclulate average / post elo and filter columns
eloFinal <- eloFinal %>%
    group_by(id) %>%
      mutate(
          op_avg_rating = round(mean(pre_rating.op),0), 
          num_games = n(),
          elo_post = round((win_elo + loss_elo) / n(),0) 
      ) %>%
    select(id , name, score, state, pre_rating, post_rating, op_avg_rating, num_games, elo_post) %>%
    distinct()

```




# Write Files

```{r, write results to file}

# write the results out to disk
write.csv(eloFinal, "tournamentinfo.csv", row.names=FALSE)

```



# Plot

```{r, plot results}

# write the results out to disk
ggplot(data=eloFinal, aes(x=name, y=op_avg_rating)) + geom_bar(stat="identity") + coord_flip()

```




