---
title: "DATA607 Project 1 - Elo "
author: "David Simbandumwe"
output:
  pdf_document: default
  html_document:
    includes:
      in_header: header.html
    css: lab.css
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float: true
editor_options: 
  
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readtext)
```


# Summary
This project starts with a text file of chess tournament results. This file is processed to develop an R Markown file that generates a .CSV file that includes:
+ Player’s Name, 
+ Player’s State, 
+ Total Number of Points, 
+ Player’s Pre-Rating, 
+ Average Pre Chess Rating of Opponent



# Load Files

```{r, read file}

 eloTxt <- read_delim( file = "https://raw.githubusercontent.com/dsimband/DATA607/main/Project1/tournamentinfo.txt",              
          delim = " ",
          col_names = c("rec"),
          trim_ws = TRUE)


# clean data frame remove extra lines
eloTxt <- eloTxt %>% 
  filter(!str_detect(rec,"\\--"))

# remove heading rows
eloTxt <- eloTxt %>% slice(-c(1,2))


```




# Process Odd Rows

```{r, process odd rows}

eloTxt_odd <- eloTxt %>% filter(row_number() %% 2 == 1)


eloTxt_odd <- eloTxt_odd %>% 
      separate(rec, 
            into = c("id", "name" , "score" , "r1", "r2", "r3", "r4", "r5", "r6", "r7","x" ),
            extra = "merge", 
            fill = "left",
            convert = TRUE,
            sep = "\\|"
            ) %>% gather(
              "r1","r2","r3","r4","r5","r6","r7", key = "round", value = "op_data"
      )%>% 
      separate(op_data, 
            into = c("r_result", "op_id"),
            convert = TRUE
      ) %>% 
      drop_na(op_id)

```



# Process Even Rows


```{r, even rows}

eloTxt_even <- eloTxt %>% filter(row_number() %% 2 == 0)


eloTxt_even <- eloTxt_even %>% 
        separate(rec, 
              into = c("state", "score_data" , "x" ),
              extra = "merge", 
              fill = "left",
              convert = TRUE,
              sep = "\\|") %>% 
        mutate(id = row_number()) %>% 
        separate(score_data, 
              into = c("x1", "x2", "x3", "pre_rating", "post_rating"),
              extra = "merge", 
              fill = "left",
              convert = TRUE) %>% 
        separate(pre_rating, 
              into = c("pre_rating", "x4"),
              extra = "merge", 
              fill = "right",
              convert = TRUE,
              sep = "P") %>% 
        separate(post_rating, 
              into = c("post_rating", "x5"),
              extra = "merge", 
              fill = "right",
              convert = TRUE,
              sep = "P") %>% 
        select(id,state,pre_rating, post_rating)


```



# Join All Rows

```{r, join odd and even rows}

eloTxt_comb  <- eloTxt_odd %>% 
  select (id, name, score, round, r_result, op_id) %>%
  left_join(eloTxt_even, by = "id") %>%
  left_join(eloTxt_even, by = c("op_id" = "id"), suffix = c("", ".op"))


```





# Calculate 

```{r, calculate elo for wins and losses}

win_df <- eloTxt_comb %>%
  group_by(id,r_result) %>%
  filter(str_detect(r_result, "W")) %>%
  mutate(
    win_elo = sum(pre_rating.op) + n()*400,
    win_num = n()
  ) %>%
  select(id,r_result,win_elo, win_num) %>%
  distinct()

loss_df <- eloTxt_comb %>% group_by(id,r_result) %>%
  filter(str_detect(r_result, "L")) %>%
  mutate(
    loss_elo = sum(pre_rating.op) - n()*400,
    loss_num = n()
  ) %>%
  select(id,r_result,loss_elo, loss_num) %>%
  distinct()

```



# create the summary table

```{r, summary table}

eloFinal <- eloTxt_comb %>% 
  select (id, name, state, score, round, r_result, op_id, pre_rating, post_rating, pre_rating.op, post_rating.op) %>%
  left_join(win_df, by = "id") %>%
  left_join(loss_df, by = "id") %>%
  replace_na(list(win_elo = 0, win_num = 0, loss_elo = 0, loss_num = 0 ))

# caclulate average / post elo and filter columns
eloFinal <- eloFinal %>%
    group_by(id) %>%
      mutate(
          op_avg_rating = round(mean(pre_rating.op),0), 
          num_games = n(),
          elo_post = round((win_elo + loss_elo) / n(),0) 
      ) %>%
    select(id , name, score, state, pre_rating, post_rating, op_avg_rating, num_games, elo_post) %>%
    distinct() %>%
    arrange(id)

```




# Write Files
The instructions say 

| When you write your .CSV, do not include a relative path (e.g. “tournament.csv” instead of "C:/Users/gerso/Documents/DATA-607/tournament.csv"), so that your code is more reproducible. (2 points)

this seems a little counter intuitive as I would assume that not everyone would have the same directory structure on their local machine. I based the code on a relative path but i also included the full path per the instructions. 

```{r, write results to file}

# write the results out to disk
write.csv(eloFinal, "tournamentinfo.csv", row.names=FALSE)

# full path
write.csv(eloFinal, "/Users/dsimbandumwe/dev/cuny/data_607/DATA607/Project1/tournamentinfo.csv", row.names=FALSE)


```



# Results
Reviewing the results of the tournament there are several big movers in both the positive and the negative directions.  The largest positive jump can be attributed to Jacob not having a high ELO rating at the start of the tournament but managing to win games against higher ranked players. The negative movement were much smaller and were caused by losses to players with lower ELO rankings


```{r, plot results}

eloFinal %>%
  mutate(
    dif_rate = post_rating - elo_post,
    post_pre = post_rating - pre_rating
  ) %>%
filter (post_pre > 50 | post_pre < -50) %>%
ggplot(aes(x=name, y=post_pre)) + geom_bar(stat="identity") + coord_flip()
  
  
eloFinal %>%
mutate(
  dif_rate = post_rating - elo_post,
  post_pre = post_rating - pre_rating
) %>%
filter (post_pre > 50 | post_pre < -50) %>%
ggplot(aes(x=name, y=dif_rate)) + geom_bar(stat="identity") + coord_flip()


```



### 46 JACOB ALEXANDER LAVALLEY
Jacob started with a low ELO rating and he competed against players with much higher ratings and he was able to defeat several players. The average ELO rating of his opponent was 1358. He had a solid tournament if you calculated his ELO rating based on his tournament performance and the average pre tournament ratings of his competitors his ELO rating would have been 1301 (ignoring his pre tournament ELO rating). 

```{r, plot 46 JACOB ALEXANDER LAVALLEY}

eloTxt_comb %>%
  left_join(eloFinal, by = c("op_id" = "id")) %>%
  filter (id == 46) %>%
  select(id, name.x, pre_rating.op, r_result, name.y) %>%
  distinct() %>%
ggplot(aes(y=name.y, x=pre_rating.op, color = r_result)) + geom_point()
  

```



### 3 ADITYA BAJAJ
Abitya performed well against players with higher rankings that accounted for the movement in his ELO score. He had a solid tournament if you calculated his ELO rating based on his tournament performance and the average pre tournament ratings of his competitors his ELO rating would have been 1849 (ignoring his pre tournament ELO rating) 

```{r, plot 3 ADITYA BAJAJ}

eloTxt_comb %>%
  left_join(eloFinal, by = c("op_id" = "id")) %>%
  filter (id == 3) %>%
  select(id, name.x, pre_rating.op, r_result, name.y) %>%
  distinct() %>%
ggplot(aes(y=name.y, x=pre_rating.op, color = r_result)) + geom_point()
  

```



###  29 CHIEDOZIE OKORIE 
Chiedoxie lost 2 games.He had a poor tournament if you calculated his ELO rating based on his tournament performance and the average pre tournament ratings of his competitors his ELO rating would have been 1099 (ignoring his pre tournament ELO rating)  

```{r, plot 29 CHIEDOZIE OKORIE}

eloTxt_comb %>%
  left_join(eloFinal, by = c("op_id" = "id")) %>%
  filter (id == 29) %>%
  select(id, name.x, pre_rating.op, r_result, name.y) %>%
  distinct() %>%
ggplot(aes(y=name.y, x=pre_rating.op, color = r_result)) + geom_point()
  

```



###  30 GEORGE AVERY JONES 
George lost 2 games to players of with lower ELO rankings. He had a poor tournament if you calculated his ELO rating based on his tournament performance and the average pre tournament ratings of his competitors his ELO rating would have been 978 (ignoring his pre tournament ELO rating)  

```{r, plot 30 GEORGE AVERY JONES }

eloTxt_comb %>%
  left_join(eloFinal, by = c("op_id" = "id")) %>%
  filter (id == 30) %>%
  select(id, name.x, pre_rating.op, r_result, name.y) %>%
  distinct() %>%
ggplot(aes(y=name.y, x=pre_rating.op, color = r_result)) + geom_point()
  

```



