---
title: "DATA607 Week 3 Assignement: R Character Manipulation and Date Processing "
author: "David Simbandumwe"
date: "11 September 2021"
output:
  prettydoc::html_pretty:
    theme: architect
    highlight: github
    
---


```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(stringr)
```





## Questions 1

**1. Using the 173 majors listed in fivethirtyeight.com’s College Majors dataset [https://fivethirtyeight.com/features/the-economic-guide-to-picking-a-college-major/], provide code that identifies the majors that contain either "DATA" or "STATISTICS"**

  
```{r}

majorsList <- read.csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/college-majors/majors-list.csv")

id = c(grep("DATA", majorsList$Major),grep("STATISTICS", majorsList$Major))

majorsList <- majorsList %>%
    mutate(search_flag = FALSE)
majorsList[id,4] = TRUE

filter(majorsList, search_flag == TRUE)

```
  
  
  
## Questions 2

**2 Write code that transforms the data below:

[1] "bell pepper"  "bilberry"     "blackberry"   "blood orange"
[5] "blueberry"    "cantaloupe"   "chili pepper" "cloudberry"  
[9] "elderberry"   "lime"         "lychee"       "mulberry"    
[13] "olive"        "salal berry"


Into a format like this:

c("bell pepper", "bilberry", "blackberry", "blood orange", "blueberry", "cantaloupe", "chili pepper", "cloudberry", "elderberry", "lime", "lychee", "mulberry", "olive", "salal berry")**

  
```{r}

str <- '[1] "bell pepper"  "bilberry"     "blackberry"   "blood orange"
[5] "blueberry"    "cantaloupe"   "chili pepper" "cloudberry"  
[9] "elderberry"   "lime"         "lychee"       "mulberry"    
[13] "olive"        "salal berry"'

writeLines(str)

str <- str_replace_all(str,"[\\[\\]]","")
str <- str_replace_all(str,"[\\n]","")
str <- str_replace_all(str,"[\\d]","")
str <- str_replace_all(str,"\\s","")
str <- str_replace_all(str,'""','","')

str <- str_c("c(",str,")")
writeLines(str)


```


## Question 3
  
**3 Describe, in words, what these expressions will match:**

**(.)\1\1**                 - (xxx) this will match to any string with 3 repeating characters 

**"(.)(.)\\2\\1"**          - (xyyx) this will match to a string that follows this patter, 1st(x) char, 2 instances of the 2nd char(yy), followed by the 1st character (x)

**(..)\1 **                 - (xyxy) this will match to a sting that follows the pattern 1st character = 3rd character and 2nd char = th character

**"(.).\\1.\\1"**           - (xyxxx) this will match to a sting that follows the pattern 1st character = 3rd through 5th charter

**"(.)(.)(.).*\\3\\2\\1" **     - (xyz\*zyx) this will match to a string that follows that pattern 3 characters, followed by one or more other characters, followed by 3rd character, the 2nd character and the 3rd character

 
## Question 4
**4 Construct regular expressions to match words that:**

Start and end with the same character. 

**- "^(.).*\\1$"**

Contain a repeated pair of letters (e.g. "church" contains "ch" repeated twice.)

**- "^(.)(.).*\\1\\2$"**

Contain one letter repeated in at least three places (e.g. "eleven" contains three "e"s.)
 
**- "(.).+\\1.+\\1"** 
  
  
This document is available at [[**RPubs**]](https://rpubs.com/dsimbandumwe/807980)
and on [[**Github**]](https://github.com/dsimband/DATA607/blob/4603e72a1ed3e5759430888d398bb2db440e27a6/Homework/DATA607_w3_DataProcessing.Rmd)
