---
title: "DATA 607 Final Project: Neural Network"
author: "David Simbandumwe"
output:
  prettydoc::html_pretty:
    theme: architect
    highlight: github
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


library(tidyverse)
library(datasets)
library(neuralnet)
library(NeuralNetTools)
library(ModelMetrics)
library(caret)
library(tidymodels)
library(dplyr)
library(plyr)

library(readr)
library(stringr)
library(glue)
library(purrr)

library(visdat)
library(GGally)
library(vip)


```


# Overview

Applying a neural network to the survey data using the same variables from the linear regression models.





## Setup

```{r, results="hide"}

set.seed(1234)
rm(list=ls())

```



```{r, include=FALSE, error=TRUE}

fun_file_name <- glue(getwd(), "/ProjectFinal/DATA607_Functions.R")
source(fun_file_name)

```




# Consumer Financial Protection Bureau Survey



## Download and Tidy dataset from CFPB

```{r}



# get cfpb file
cfpb_df <- getCFPBFile()
cfpb_df$cfpb_score_4cat <- cut(cfpb_df$cfpb_score, breaks = c(-10, 40, 60, 80, 100),
                           labels = c("< 40","40-60","60-80","80-100"),
                           right = FALSE,
                           include.lowest=TRUE) 
cfpb_df <- cfpb_df %>% filter(cfpb_score >= 0)

# reduce cfpb data set
cfpb_df <- slice_sample(cfpb_df, weight_by=cfpb_score_4cat ,n=5000) 
cfpb_df <- cfpb_df %>% select(cfpb_score, econ_save_rate, house_mortgage, age_8cat, econ_hh_income)

summary(cfpb_df$cfpb_score)


# prepare data for analysis
concrete_norm <- cfpb_df
concrete_norm$cfpb_score <- normalize(concrete_norm$cfpb_score)

dmy <- dummyVars(" ~ .", data = concrete_norm, fullRank = T)
dat_transformed <- data.frame(predict(dmy, newdata = concrete_norm))
glimpse(dat_transformed)
summary(concrete_norm$cfpb_score)


# split data
split <- initial_split(dat_transformed, prop = .8, strata = cfpb_score)

train_data <- training(split) 
test_data <- testing(split)


```



## Train Model

```{r}

# create model
nn <- neuralnet(cfpb_score ~ ., data = train_data, rep=1 , stepmax=1e5, err.fct="sse")
nn$result.matrix
plot(nn)

```



## Test Model

```{r}

len <- length(names(test_data))

results <- compute(nn, test_data[2:len])

predicted_score <- results$net.resul


```


```{r}




t1 <- tibble(
    predict = results$net.result,
    actual = test_data$cfpb_score
)
t1 <- t1 %>% mutate (diff = predict - actual)

summary(t1)

ggplot(data = t1,
       mapping = aes(x = predict, y = actual)) +
  geom_point(color = '#006EA1') +
  geom_abline(intercept = 0, slope = 1, color = 'orange') +    
  labs(title = 'Neural Net (cfpb data set) ',
       x = 'Predicted',
       y = 'Actual')


ggplot(data = t1) +
  geom_boxplot(mapping = aes(x=diff)) +
  labs (title = "Std Error - Neural Net (cfpb data set)")

ggplot(t1,
    aes(diff, y = stat(density))) +
    geom_histogram(binwidth = .02, alpha = 0.7, bins = 100, color="white",size = 0.1) +
    geom_vline(aes(xintercept = mean(diff)), linetype = "dashed", size = 0.8, color="red") +
    scale_fill_brewer(palette="Spectral") +
    labs(title = "Std Error - Neural Net (cfpb data set)")


```





### Adding 5 Hidden layers to improve model performance.

```{r}

nn2 <- neuralnet(cfpb_score ~ ., data = train_data, rep=1 , stepmax=1e5, hidden = 5)
nn$result.matrix
plot(nn2)

```



```{r}

len <- length(names(test_data))

results2 <- compute(nn2, test_data[2:len])

predicted_score2 <- results2$net.result


```



```{r}


t2 <- tibble(
    predict = results2$net.result,
    actual = test_data$cfpb_score
)
t2 <- t2 %>% mutate (diff = predict - actual)

summary(t2)

ggplot(data = t2,
       mapping = aes(x = predict, y = actual)) +
  geom_point(color = '#006EA1') +
  geom_abline(intercept = 0, slope = 1, color = 'orange') +    
  labs(title = 'Neural Net - 5 hidden layers (cfpb data set)',
       x = 'Predicted',
       y = 'Actual')

ggplot(data = t2) +
  geom_boxplot(mapping = aes(x=diff)) +
  labs (title = "Std Error - Neural Net (cfpb data set)")

ggplot(t2,
    aes(diff, y = stat(density))) +
    geom_histogram(binwidth = .02, alpha = 0.7, bins = 100, color="white",size = 0.1) +
    geom_vline(aes(xintercept = mean(diff)), linetype = "dashed", size = 0.8, color="red") +
    scale_fill_brewer(palette="Spectral") +
    labs(title = "Std Error - Neural Net (cfpb data set)")

```











# Federal Reserve System Survey

## Download and Tidy dataset from FED

```{r}


# get fed file
fed_df <- getFedFile()

#filter data
fed_df <- fed_df %>% drop_na()
fed_df <- fed_df %>% filter(!(econ_saving == "" |  credit_guess =="" | health ==""))
fed_df$credit_guess <- str_replace_all(fed_df$credit_guess, "[^[:alnum:]]", "")


fed_df$cfpb_score_4cat <- cut(fed_df$cfpb_score, breaks = c(-10, 40, 60, 80, 100),
                           labels = c("< 40","40-60","60-80","80-100"),
                           right = FALSE,
                           include.lowest=TRUE) 

# reduce cfpb dataset
fed_df <- slice_sample(fed_df, weight_by=cfpb_score_4cat ,n=5000) 

fed_df <- fed_df %>% select(cfpb_score, age_7cat, econ_saving, econ_inc_4cat, econ_fin_ok, 
                            econ_pay_exp400, econ_skip_med)

summary(fed_df$cfpb_score)

# prepare data for analysis
concrete_norm <- fed_df
concrete_norm$age <- normalize(as.numeric(concrete_norm$age))
concrete_norm$cfpb_score <- normalize(as.numeric(concrete_norm$cfpb_score)) 


dmy <- dummyVars(" ~ .", data = concrete_norm, fullRank = T)
dat_transformed <- data.frame(predict(dmy, newdata = concrete_norm))
glimpse(dat_transformed)
summary(dat_transformed$cfpb_score)


# split data
split <- initial_split(dat_transformed, prop = .8, strata = cfpb_score)

train_data <- training(split) 
test_data <- testing(split)


```

## Train Model

```{r}

# create model
nn <- neuralnet(cfpb_score ~ ., data = train_data, rep=1 , stepmax=1e5, err.fct="sse")
nn$result.matrix
plot(nn)


```



## Test Model

```{r}

len <- length(names(train_data))

results <- compute(nn, test_data[2:len])

predicted_score <- results$net.result


```

```{r}

t1 <- tibble(
    predict = results$net.result,
    actual = test_data$cfpb_score
)
t1 <- t1 %>% mutate (diff = predict - actual)

summary(t1)


ggplot(data = t1,
       mapping = aes(x = predict, y = actual)) +
  geom_point(color = '#006EA1') +
  geom_abline(intercept = 0, slope = 1, color = 'orange') +    
  labs(title = 'Std Error - Neural Net (FED data set)',
       x = 'Predicted',
       y = 'Actual')


ggplot(data = t1) +
  geom_boxplot(mapping = aes(x=diff)) +
  labs (title = "Std Error - Neural Net (FED data set)")

ggplot(t1,
    aes(diff, y = stat(density))) +
    geom_histogram(binwidth = .02, alpha = 0.7, bins = 100, color="white",size = 0.1) +
    geom_vline(aes(xintercept = mean(diff)), linetype = "dashed", size = 0.8, color="red") +
    scale_fill_brewer(palette="Spectral") +
    labs(title = "Std Error - Neural Net (FED data set)")

```





## Adding 5 Hidden layers to improve model performance.

```{r}

nn2 <- neuralnet(cfpb_score ~ ., data = train_data, rep=1 , stepmax=1e5, hidden = 5)
nn$result.matrix
plot(nn2)

```



```{r}

len <- length(names(train_data))

results2 <- compute(nn2, test_data[2:len])

predicted_score2 <- results2$net.result



```


```{r}
t2 <- tibble(
    predict = results2$net.result,
    actual = test_data$cfpb_score
)
t2 <- t2 %>% mutate (  
    diff = actual - predict,
    dev = (actual - predict) / actual
    )
#accuracy_res = 1-abs(mean(t2$dev))
summary(t2)



ggplot(data = t2,
       mapping = aes(x = predict, y = actual)) +
  geom_point(color = '#006EA1') +
  geom_abline(intercept = 0, slope = 1, color = 'orange') +    
  labs(title = 'Neural Net 5 layers (FED data set) ',
       x = 'Predicted',
       y = 'Actual')


ggplot(data = t2) +
  geom_boxplot(mapping = aes(x=diff)) +
  labs (title = "Std Error - Neural Net (FED data set")

ggplot(t2,
    aes(diff, y = stat(density))) +
    geom_histogram(binwidth = .02, alpha = 0.7, bins = 100, color="white",size = 0.1) +
    geom_vline(aes(xintercept = mean(diff)), linetype = "dashed", size = 0.8, color="red") +
    scale_fill_brewer(palette="Spectral") +
    labs(title = "Std Error - Neural Net (FED data set)")


```



# Conclusion
	
Developing the neural net was an interesting exercise however it might not be the most appropriate modeling approach for the project. 
	
	
	
	
	
	
