---
title: "DATA 607 Final Project: Model Selection"
author: "David Simbandumwe"
output:
  prettydoc::html_pretty:
    theme: architect
    highlight: github
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(utils)

library(tidyverse)
library(tidymodels)
library(textrecipes)
library(discrim)
library(dplyr)
library(plyr)
library(naivebayes)

library(glue)
library(vip)

library(GGally)

```


# Overview
Stepwise model selection process

## Setup

```{r, results="hide"}

set.seed(1234)
rm(list=ls())

```


```{r, include=FALSE, error=TRUE}

fun_file_name <- glue(getwd(), "/ProjectFinal/DATA607_Functions.R")
source(fun_file_name)


```



# Consumer Financial Protection Bureau


## Download and Tidy dataset from CFPB

```{r}

cfpb_df <- getCFPBFile()
cfpb_df$cfpb_score_4cat <- cut(cfpb_df$cfpb_score, breaks = c(-10, 40, 60, 80, 100),
                           labels = c("< 40","40-60","60-80","80-100"),
                           right = FALSE,
                           include.lowest=TRUE) 

cfpb_df <- cfpb_df %>% filter(cfpb_score >= 0)
cfpb_df <- slice_sample(cfpb_df, weight_by=cfpb_score_4cat ,n=4000) 

# remove calculated variables 
cfpb_df <- cfpb_df %>% select(-c(cfpb_score_4cat, fs_skills_score, fs_lm_score, fs_kh_score))
cfpb_base_df <- cfpb_df


# select independent variables
#cfpb_df <- cfpb_df %>% select(cfpb_score, econ_hh_income, econ_save_rate, age_8cat) # 0.3984 

# split data
cfpb_split <- initial_split(cfpb_df, prop = 0.8, strata = cfpb_score)
cfpb_training <-  training(cfpb_split)
cfpb_test <-  testing(cfpb_split)

```

## full model

```{r}

cfpb_result1<-lm(cfpb_score ~ . ,data=cfpb_training)
summary(cfpb_result1)

```





## Stepwise regression both

The function step should be called and the direction is set to both so that the algorithm can add and drop predictors in every iteration. Finally, when dropping/adding any variable will not give a positive impact to our model in terms of performance, the stepwise process is done


```{r}

cfpb_results2<-stats::step(cfpb_result1, direction = "both")

```


## Stepwise regression Akaike information criterion (AIC)

For AIC and BIC, we use step function again and specify the penalty for the number of freedom. For instance, we have k=2
 in AIC and k=log(n) in BIC, implying that BIC normally gives a larger penalty on the number of parameters in the model according to the definition of BIC. Once the model selection is done, we should expect a smaller number of predictors in BIC than AIC below.


```{r}
cfpb_result3<- stats::step(cfpb_result1, k=2)
```

## Bayes information criterion (BIC)


```{r}

cfpb_result4<-stats::step(cfpb_result1, k=log(nrow(cfpb_training)))

```


## explore both

```{r}

# Step:  AIC=15031.65
# cfpb_score ~ sample_pop + econ_save_rate + house_value + house_mortgage + 
#     age_8cat + race + econ_hh_income + region_9base

cfpb_result_c1 <-lm(cfpb_score ~ sample_pop + econ_save_rate + house_value + house_mortgage + 
    age_8cat + race + econ_hh_income + region_9base,
     data=cfpb_training)
summary(cfpb_result_c1)

```


## explore Bayes information criterion (BIC)


```{r}

# Step:  AIC=18967.84
# cfpb_score ~ econ_save_rate + house_mortgage + age_8cat + econ_hh_income 

cfpb_result_c2 <-lm(cfpb_score ~ econ_save_rate + house_mortgage + age_8cat + econ_hh_income, data=cfpb_training)
summary(cfpb_result_c2)


```








# Federal Reserve





## Download and Tidy dataset from Fed

```{r}


# get fed file
fed_df <- getFedFile()
fed_df$cfpb_score_4cat <- cut(fed_df$cfpb_score, breaks = c(-10, 40, 60, 80, 100),
                           labels = c("< 40","40-60","60-80","80-100"),
                           right = FALSE,
                           include.lowest=TRUE) 


# reduce cfpb dataset
fed_df <- slice_sample(fed_df, weight_by=cfpb_score_4cat ,n=4000) 


# remove categorical variable for data set
fed_df <- fed_df %>% select(-c(cfpb_score_4cat))

# split data
fed_split <- initial_split(fed_df, prop = 0.8, strata = cfpb_score)
fed_training <- training(fed_split)
fed_test <- testing(fed_split)



```





## full model

```{r}

fed_result1<-lm(cfpb_score ~ . ,data=fed_training)
summary(fed_result1)


```




## Stepwise regression both

```{r}

fed_results2<-stats::step(fed_result1, direction = "both")

```





## Stepwise regression Akaike information criterion (AIC)

```{r}

fed_result3<- stats::step(fed_result1, k=2)

```




## Bayes information criterion (BIC)


```{r}

fed_result4<-stats::step(fed_result1, k=log(nrow(fed_training)))

```




## explore both

```{r}

# Step:  AIC=14863.85
# cfpb_score ~ gender + age + age_7cat + household_size + hh_child_0.1 + 
#     hh_child_13.17 + hh_child_18p + emp_status + emp_union + 
#     health + emp_type + race + edu_cat + race_census + credit_guess + 
#     econ_saving + econ_inc_4cat + econ_fin_ok + econ_pay_exp400 + 
#     econ_skip_med

fed_result_c1 <-lm(cfpb_score ~ gender + age + age_7cat + household_size + hh_child_0.1 + 
    hh_child_13.17 + hh_child_18p + emp_status + emp_union + 
    health + emp_type + race + edu_cat + race_census + credit_guess + 
    econ_saving + econ_inc_4cat + econ_fin_ok + econ_pay_exp400 + 
    econ_skip_med,
     data=fed_training)
summary(fed_result_c1)

```



## explore Bayes information criterion (BIC)

```{r}

# Step:  AIC=15103.63
# cfpb_score ~ age + age_7cat + econ_saving + econ_inc_4cat + econ_fin_ok + 
#     econ_pay_exp400 + econ_skip_med


fed_result_c2 <-lm(cfpb_score ~ age + age_7cat + econ_saving + econ_inc_4cat + econ_fin_ok + 
    econ_pay_exp400 + econ_skip_med,
     data=fed_training)
summary(fed_result_c2)


```



## manual mapping

```{r}

# Step: manual forwards regression
# cfpb_score ~ econ_saving + econ_fin_ok + econ_pay_exp400 + econ_skip_med + edu + credit_guess

fed_result_c3 <-lm(cfpb_score ~ econ_saving + econ_fin_ok + econ_pay_exp400 + 
                       econ_skip_med + edu + credit_guess,
     data=fed_training)
summary(fed_result_c3)



```



```{r}

# Step:  AIC=15103.63
# cfpb_score ~ age + econ_saving + econ_inc_4cat + econ_fin_ok + 
#     econ_pay_exp400 + econ_skip_med


fed_result_c2 <-lm(cfpb_score ~ age + econ_saving + econ_inc_4cat + econ_fin_ok + 
    econ_pay_exp400 + econ_skip_med,
     data=fed_training)
summary(fed_result_c2)

```




```{r}

# Step:  AIC=15103.63
# cfpb_score ~ age_7cat + econ_saving + econ_inc_4cat + econ_fin_ok + 
#     econ_pay_exp400 + econ_skip_med


fed_result_c2 <-lm(cfpb_score ~ age_7cat + econ_saving + econ_inc_4cat + econ_fin_ok + 
    econ_pay_exp400 + econ_skip_med,
     data=fed_training)
summary(fed_result_c2)

```




# final models - CFPB

```{r}

# Step:  AIC=18967.84
# cfpb_score ~ econ_save_rate + house_mortgage + age_8cat + econ_hh_income 

cfpb_result_c2 <-lm(cfpb_score ~ econ_save_rate + house_mortgage + age_8cat + econ_hh_income, data=cfpb_training)
summary(cfpb_result_c2)


```


# final model - FED

```{r}

# Step:  AIC=15103.63
# cfpb_score ~ age_7cat + econ_saving + econ_inc_4cat + econ_fin_ok + 
#     econ_pay_exp400 + econ_skip_med


fed_result_c2 <-lm(cfpb_score ~ age_7cat + econ_saving + econ_inc_4cat + econ_fin_ok + 
    econ_pay_exp400 + econ_skip_med,
     data=fed_training)
summary(fed_result_c2)

```


